Macroscopic Salt Rejection through Electrostatically Gated Nanoporous Graphene

### 2019-03-06 paper/SI.tex
Add more analysis into SI

\documentclass[manuscript=suppinfo,email=true, hyperref=true, keywords=false]{achemso}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{amsmath}
\usepackage{times}
\usepackage{todonotes}
\newcommand{\Fig}{Fig.}
\renewcommand{\thesection}{S\arabic{section}}
\renewcommand{\theequation}{S\arabic{equation}}
\SectionNumbersOn
\usepackage{xr-hyper}
\externaldocument[main-]{paper}

@@

\pagebreak
\section{Further details of the numerical simulation }
\label{sec:numer}

@@

\section{Further numerical simulation results}
\label{sec:simu-res}

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.8\linewidth]{img/SI-concentration.png}
  \caption{Cation concentration $c_{+}$ (\textbf{a}.) and anion
    concentration $c_{-}$ (\textbf{b}.) near the graphene nanopore for
    KCl solution at 0.1 mM. Depletion of cation and accumulation of
    anion can be observed near the graphene surface.}
  \label{fig:conc}
\end{figure}

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.8\linewidth]{img/SI-electrochemical-decomposite.pdf}
  \caption{Decomposition of electrochemical potential near a nanopore
    corresponding to \Fig 4 in main text. \textbf{a}. The electrostatic
    (drift) contribution to the electrochemical potential for cation
    (left) and anion (right) and \textbf{b}. The concentration (diffusion)
    contribution to the electrochemical potential for cation (left)
    and anion (right). The relative small order of magnitude for anion
    electrochemical potential is caused by the balance between the
    diffusion and drift for anion.}
  \label{fig:potential}
\end{figure}

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.8\linewidth]{img/SI-flux.png}
  \caption{z-component flux $\boldsymbol{J}_{z}$ of cation
    (\textbf{a}.)  and anion (\textbf{b}.) near the graphene nanopore
    corresponding to Figure 4b in main text. The pathways for the
    different ions can be spatially distinguished: the flux of cation
    passes mostly through the center of the graphene nanopore while
    the anion passes mainly through the edge. The analysis is in
    consistent with the gradient of chemical potential as shown in the
    main text \Fig 4b.}
  \label{fig:flux}
\end{figure}

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.5\linewidth]{img/SI-rect-reverse.pdf}
  \caption{Average total flux $\boldsymbol{J}_{z}$ in the nanopore as
    a function of $\psi_{\mathrm{G}}$ at different concentrations. The
    range of $\psi_{\mathrm{G}}$ is larger than the experimentally
    achievable value. The rectification of flux becomes weaker after a
    certain level of $\psi_{\mathrm{G}}$ is reached. Further
    increasing the surface potential of graphene may even enhance the
    transportation of ions, which is similar to the cases of ionic
    transistors.}
  \label{fig:reverse}
\end{figure}

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.8\linewidth]{img/SI-flux-larger.png}
  \caption{Spatial distribution of $\boldsymbol{J}_{z}$ as a function
    of $r$ in a 20 nm nanopore at different levels of
    $\psi_{\mathrm{G}}$ (larger than experimentally achievable values)
    for cation (\textbf{a}.) and anion (\textbf{b}.). The cation flux
    becomes nearly saturated with increasing $\psi_{\mathrm{G}}$,
    while the flow of anion near the pore edge becomes much
    larger. The drastic increase of pore-bounded anion flux is caused
    by the accumulation of anions near the graphene surface, and in
    turn causes the increasing of total flux as seen in \Fig
    \ref{fig:reverse}.}
  \label{fig:large-V}
\end{figure}
\clearpage{}

>>>>

\documentclass[manuscript=suppinfo,email=true, hyperref=true, keywords=false]{achemso}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{amsmath}
\usepackage{times}
\usepackage{todonotes}
\newcommand{\Fig}{Fig.}
\renewcommand{\thesection}{S\arabic{section}}
\renewcommand{\theequation}{S\arabic{equation}}
\SectionNumbersOn
\usepackage{xr-hyper}
\usepackage{xcolor}
\externaldocument[main-]{paper}

@@

\section{Effect of pore size distribution}
\label{sec:pore-dist}

One practical issue with large-area nanoporous graphene samples is
that the unavoidable large pores will act as preferred pathways for
the transported species and strongly attenuates the salt rejection
effect. In this section we provide a detailed discussion about the
influence of pore size distribution on the salt rejection through the
PG-PCTE membrane. As seen from main text \Fig{} \ref{main-fig:1}c
inset, the average distance between individual nanopores is larger
than the largest Debye length studied here ($\sim{}$ 30 nm for 1 mM 1:1
salt solution), we can assume that the ionic transport pathways
through different nanopores do not influence each other and the total
ionic flux $\boldsymbol{J}_{\mathrm{PG}}$ is the summation of fluxes
through individual pores $\boldsymbol{J}_{i}$:
\begin{equation}
\label{eq:J-total}
\boldsymbol{J}_{\mathrm{PG}} = \sum_{i} \boldsymbol{J}_{i} = \sum_{r_{i}} x_{r_{i}} \boldsymbol{J}_{r_{i}}
\end{equation}
where $r_{i}$ is the radius of individual pores,
$\boldsymbol{J}_{r_{i}}$ and $x_{r_{i}}$ are the flux of individual
pore and the distribution probability when
$r_{\mathrm{G}} = r_{i}$. The salt rejection factor
$\xi$ through graphene is then expressed as:
\begin{equation}
  \label{eq:xi-pore}
  \begin{aligned}
    \xi &= 1 - \frac{\sum_{r_{i}} \boldsymbol{J}_{r_{i}} x_{r_{i}}}
           {\sum_{r_{i}} \boldsymbol{J}_{r_{i}}(V_{\mathrm{G}} = 0) x_{r_{i}}} \\
           &= 1 - \frac{\sum_{r_{i}} \hat{J}_{r_{i}} r_{i}^{2} x_{r_{i}}}
           {\sum_{r_{i}} \hat{J}_{r_{i}}^{0} r_{i}^{2} x_{r_{i}}} \\
           % &approx 1 - \sum_{r_{i}} \frac{\hat{J}_{r_{i}}}{\hat{J}_{r_{i}}^{0}}
           % \frac{r_{i}^{2} x_{r_{i}}} {\sum_{r_{i}} r_{i}^{2} x_{r_{i}}} \\
           &\approx \sum_{r_{i}} \left(1 -  \frac{\hat{J}_{r_{i}}}{\hat{J}_{r_{i}}^{0}}\right) w_{r_{i}} \\
           &= \sum_{r_{i}} \xi_{r_{i}} w_{r_{i}}
  \end{aligned}
\end{equation}
where $\hat{J}$ and $\hat{J}^{0}$ are the flux normalized by pore area
when $V_{\mathrm{G}} = 0$ and $V_{\mathrm{G}} \neq 0$, respectively,
$w_{r_{i}} = x_{r_{i}} r_{i}^{2} / sum_{r_{i}} x_{r_{i}} r_{i}^{2}$ is
the contribution to the salt rejection from individual pore rejection
rate $\xi_{r_{i}}$. The linear combination between $\xi_{r_{i}}$ and
$\xi$ comes from the assumption that $\hat{J}^{0}$ is independent of
the pore size from the simple Ficknian diffusion.

To get an insight of the influence of the pore size distribution on
the salt rejection, we combine Equation \ref{eq:xi-pore} with main
text \Fig{} \ref{main-fig:5}b for the salt rejection of a 1:1
electrolyte to get the pore-distribution-related salt rejection, where
the pore distribution is taken from the SEM image of real samples (for
instance main text \Fig{} \ref{main-fig:1}c inset). \Fig{}
\ref{fig:simple-rect-pore}a shows the contribution $w$ and $\xi$ as
a function of $r_{\mathrm{G}}$ for individual pores when
$V_{\mathrm{G}}$ = 1.25 V. As can be seen, the majority of pores fall
within the regime of $\lambda_{\mathrm{D}} / r_{\mathrm{G}} > 1$,
corresponding to $\xi(r_{\mathrm{G}}) > 0.6$. As a result, considering
the pore distribution reduces the value of $\xi$ to $\sim{}$80\% to
that of a single 20 nm-diameter nanopore (\Fig{}
\ref{fig:simple-rect-pore}b), which is indeed close to the highest
$\xi$ values experimentally observed (for 1mM NaCl system). Such high
salt rejection even after considering the pore size distribution, can
be explained by the non-linear nature of $\xi$ with $r_{\mathrm{G}}$
and $V_{\mathrm{G}}$. As seen in main text Fig{} \ref{main-fig:5}b,
for large $V_{\mathrm{G}}$ and $\lambda_{\mathrm{D}} / r_{\mathrm{G}}$
regions, $\xi$ reaches a ``plateau'' with its value close to 1. For
the fabricated PG samples, most pores are within such region when
$V_{\mathrm{G}}$ = 1.25 V and $c_{0}$ = 0.1 mM, which gives a minor
change of $\xi$ dispite the wide span of pore sizes. In other words,
although nanopres as large as 60 nm exist in the system, considerably
high salt rejection more than 80\% can still be achieved.

\begin{figure}[htbp]
  \centering
   \includegraphics[width=0.9\linewidth]{img/SI-simple-pore-dist.pdf}
   \caption{Salt rejection after considering the pore size
     distribution. \textbf{a}. $w$ (left axis) and $\xi$ (right axis)
     as a function of $r_{\mathrm{G}}$ for the experimental pore size
     distribution with $c_{0}$ = 0.1 mM and $V_{\mathrm{G}}$ = 1.25 V.
     \textbf{b}. Salt rejection ratio $\xi$ as a function of $V_{\mathrm{G}}$ of a
     0.1mM 1:1 salt from a 20 nm-diameter nanopore (blue line) and
     from the experimental pore size distribution (orange line), by
     applying Equation \ref{eq:xi-pore} to main text \Fig{}
     \ref{main-fig:5}b.}
  \label{fig:simple-rect-pore}
\end{figure}

To include the pore size distribution effect for all the salts
studied, we performed FEM analysis for the salt rejection through
single nanopore of different salts with $r_{\mathrm{G}}$ ranges from
2.5 nm to 35 nm, while keeping $c_{0}$ constant at 0.1 mM. As shown in
\Fig{} \ref{fig:xi-salts-pore}, the behavior of $\xi$ vary with the
type of salt. As expected, the overall $\xi$ for monovalent salts
(KCl, NaCl, LiCl) is larger than multivalent salts due to the longer
Debye length. Moreover, $\xi$ of monovalent salts varies less with the
pore size compared with multivalent salts. This can be explained by
the non-linear dependency of $\xi$ on
$\lambda_{\mathrm{D}} / r_{\mathrm{G}}$ as discussed above. Combine
the FEM-simulated $\xi$ values with the experimental pore size
distribution, we obtain main text \Fig{} \ref{main-fig:6}.

\begin{figure}[htbp]
  \centering
   \includegraphics[width=0.75\linewidth]{img/SI-xi-salt-pore-size.pdf}
  \caption{FEM-simulated salt rejction ratio $\xi$ as a function of
    $V_{\mathrm{G}}$ through single nanopores for different salts with
    varied pore radius. All salts have concentration $c_{0}$ = 0.1 mM.}
  \label{fig:xi-salts-pore}
\end{figure}

\pagebreak

\section{Further discussion about the ionic pathway near the graphene pore}
\label{sec:conc}

In this section we provide a more detailed discussion about the
different ionic pathways through the graphene nanopore upon gating in
main text \Fig{} \ref{main-fig:4}b. As disccused in the main text, the
ionic migration is contributed by both diffusion (caused by
concentration gradient) and electrostatic drift (caused by electric
field).  The effect of salt rejection can be viewed as follows: when
$V_{\mathrm{G}}$ = 0, the flux $\boldsymbol{J}_{\mathrm{G}}$ is solely
contributed by the concentration gradient, while after applying a
positive $V_{\mathrm{G}}$, distinct pathways for cations /
anions are created and the drift flux balances the diffusion.

Applying a positive $V_{\mathrm{G}}$ changes the surface distribution
of the cations / anions. To see this, we plot the concentrations of
cation ($c_{+}$) and anion ($c_{-}$) near a graphene nanopore with
$r_{\mathrm{G}}$ = 10 nm, $c_{0}$ = 0.1 mM and $\psi_{\mathrm{G}}$ =
0.15 V, as shown in \Fig{} \ref{fig:conc}a and \ref{fig:conc}b,
respectively. The cations are depleted from the graphene surface,
while the anions are accumulated due to positive
$\psi_{\mathrm{G}}$. The anion concentration at the graphene surface
is enhanced $\sim{}$35 times compared with the bulk concentration in
HCR. The thickness of the depletion / accumulation layer is affected
by the bulk concentration. Here we show this by comparing the radial
distribution of $c_{+}$ and $c_{-}$ along the $z = 0$ line within the
graphene pore. \Fig{} \ref{fig:conc-r} shows such radial ionic
concentration distribution at different $c_{0}$ and $V_{\mathrm{G}}$
levels. When $c_{0}$ = 0.1 mM, $\lambda_{\mathrm{D}}$ is larger than
$r_{\mathrm{G}}$, and the external gate voltage $V_{\mathrm{G}}$ has
control over both $c_{-}$ and $c_{+}$ within the graphene nanopore
(\Fig{} \ref{fig:conc-r}a and \ref{fig:conc-r}c). On the other
hand, when $c_{0}$ = 100 mM, $\lambda_{\mathrm{D}}$ is much smaller
than $r_{\mathrm{G}}$, and the concentration is only affected by
$V_{\mathrm{G}}$ close to the pore edge (\Fig{} \ref{fig:conc-r}b
and \ref{fig:conc-r}d). Therefore by increasing $c_{0}$ (and
equivalently reducing $\lambda_{\mathrm{D}}$), $V_{\mathrm{G}}$ loses
control over the transport in the center of the pore and leads to a
lower $\xi$ value. Due to the existence of graphene quantum
capacitance, the surface potential $\psi_{\mathrm{G}}$ is essentially
smaller than $V_{\mathrm{G}}$, and the maximum surface anion
concentration ($c_{0}$ = 100 mM, $V_{\mathrm{G}}$ = 1.25 V) is
$\sim{}$3 M, much smaller than the saturated surface adsorption
density (at the order of 10$^{2}$ M
\cite{bard_electrochemical_1980}). Therefore we expect the surface
enhancement of the anions under the FEM conditions to be realistic and
the influence of the anion concentration on the diffusivity is minimal
\cite{Tang_1999_I}.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.8\linewidth]{img/SI-concentration.pdf}
  \caption{Cation concentration $c_{+}$ (\textbf{a}.) and anion
    concentration $c_{-}$ (\textbf{b}.) near the graphene nanopore
    with $r_{\mathrm{G}}$ = 10 nm, $c_{0}$ = 0.1 mM, and
    $\psi_{\mathrm{G}}$ = 0.15 V, for a KCl solution at 0.1
    mM. Depletion of cation and accumulation of anion can be observed
    near the graphene surface.}
  \label{fig:conc}
\end{figure}

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.75\linewidth]{img/SI-conc-r.pdf}
  \caption{Radial distribution of ions along the line $z=0$ within a
    graphene nanopore with $r_{\mathrm{G}}$ = 10 nm: (\textbf{a}.)
    $c_{+}$ at $c_{0}$ = 0.1 mM, (\textbf{b}.) $c_{+}$ at $c_{0}$ =
    100 mM, (\textbf{c}.) $c_{-}$ at $c_{0}$ = 0.1 mM and
    (\textbf{d}.) $c_{-}$ at $c_{0}$ = 100 mM. The lines with varied
    colors indicate the difference of $V_{\mathrm{G}}$ applied.}
  \label{fig:conc-r}
\end{figure}

Next we analyze the diffusion and drift contributions to the ionic
transport. As shown in \Fig{} \ref{fig:potential}, the
electrochemical potential energy $\mu_{i}$ is decomposed into the
electric potential energy $z_{i} c_{i} \psi_{i}$ (\Fig{}
\ref{fig:potential}a and \ref{fig:potential}b) and the chemical
potential energy $k_{\mathrm{B}}T \ln x_{i}$ (\Fig{}
\ref{fig:potential}c and \ref{fig:potential}d) near a graphene
nanopore with $r_{\mathrm{G}}$ = 10 nm and $\psi_{\mathrm{G}}$ = 0.15
V. The transport of cations is dominated by its chemical potential and
the diffusion is prominent in the center of the pore. On the other
hand, the electric and diffusive potential energies are similar with
opposite signs and counterbalance each other. This explains the
relatively smaller electrochemical potential of anions in main text
\Fig{} \ref{main-fig:4}b. As a result, different pathways for cations
and anions emerge that cations preferably pass through the pore
center, while anions mainly pass near the pore edge, as shown in
\Fig{} \ref{fig:flux}.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.8\linewidth]{img/SI-electrochemical-decomposite.pdf}
  \caption{Decomposition of electrochemical potential near a nanopore
    corresponding to main text \Fig \ref{main-fig:4}. \textbf{a}. The electrostatic
    (drift) contribution to the electrochemical potential for cation
    (left) and anion (right) and \textbf{b}. The concentration (diffusion)
    contribution to the electrochemical potential for cation (left)
    and anion (right). The relative small order of magnitude for anion
    electrochemical potential is caused by the balance between the
    diffusion and drift for anion.}
  \label{fig:potential}
\end{figure}

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.8\linewidth]{img/SI-flux.png}
  \caption{z-component fluxes of cation ($J_{z+}$, \textbf{a}.)  and
    anion ($J_{z-}$ \textbf{b}.) near the graphene nanopore
    corresponding to \Fig{} \ref{main-fig:4}b in main text. The
    pathways for the different ions can be spatially distinguished:
    the flux of cation passes mostly through the center of the
    graphene nanopore while the anion passes mainly through the
    edge. The analysis is in consistent with the gradient of chemical
    potential as shown in the main text \Fig{} \ref{main-fig:4}b.}
  \label{fig:flux}
\end{figure}

We note that when the surface potential $\psi_{\mathrm{G}}$ of
graphene further increases, the salt rejection may be weakened, or
even enhanced salt transport is observed. To see this we artificially
increase the $\psi_{\mathrm{G}}$ up to over 1 V when not considering
the limiting effect of graphene quantum capacitance. As seen in \Fig{}
\ref{fig:reverse}, for each salt concentration $c_{0}$, the averaged
total flux $J_{z}$ does not further decrease when $\psi_{\mathrm{G}}$
is larger than a critical potential $V_{\mathrm{c}}$. At low $c_{0}$,
further increasing $\psi_{\mathrm{G}}$ over $V_{\mathrm{c}}$ may even
enhance the flux, leading to a negative $\xi$. The reserve of the salt
rejection at higher $V_{\mathrm{G}}$ values is caused by the enhanced
anion flux around the pore edge, while the cation flux saturates
(\Fig{} \ref{fig:large-V}). Such flux enhancement resemble that in an
ionic transistor \cite{Nam_2009,Cheng_2018} when ionic current rises
by applying gate voltage. Nevertheless, such regime may not be easy to
be observed experimentally as the $V_{\mathrm{G}}$ applied may already
exceed the electrochemical window of water, due to the quantum
capacitance of graphene.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.5\linewidth]{img/SI-rect-reverse.pdf}
  \caption{Average total flux $J_{z}$ in the nanopore as
    a function of $\psi_{\mathrm{G}}$ at different concentrations. The
    range of $\psi_{\mathrm{G}}$ is larger than the experimentally
    achievable value. The rectification of flux becomes weaker after a
    certain level of $\psi_{\mathrm{G}}$ is reached. Further
    increasing the surface potential of graphene may even enhance the
    transportation of ions, which is similar to the cases of ionic
    transistors.}
  \label{fig:reverse}
\end{figure}

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.8\linewidth]{img/SI-flux-larger.png}
  \caption{Spatial distribution of $J_{z}$ at $z = 0$ as a function
    of $r$ in a nanopore with $r_{\mathrm{G}}$ = 20 nm, at different
    levels of $\psi_{\mathrm{G}}$ (larger than experimentally
    achievable values) for cation (\textbf{a}.) and anion
    (\textbf{b}.). The cation flux becomes nearly saturated with
    increasing $\psi_{\mathrm{G}}$, while the flow of anion near the
    pore edge becomes much larger. The drastic increase of
    pore-bounded anion flux is caused by the accumulation of anions
    near the graphene surface, and in turn causes the increasing of
    total flux as seen in \Fig \ref{fig:reverse}.}
  \label{fig:large-V}
\end{figure}
\clearpage{}

\section{Further details of the numerical simulation }
\label{sec:numer}

@@
